Reformulating phylogenetic models to improve speed and flexibility

### 2018-04-18 main.tex
working on results

Ever-increasing capabilities of data collection techniques (i.e. genetic sequencing, geographic information systems monitoring, and etc) and sampling efforts from field biologist are making large volumes of ecological/biological data available over an ever-wider range of species and also higher resolution (i.e. measurable phenotypic traits to genetic sequence data). 
As a result, researchers are interested in exploring more complicated systems and fitting more models to data or synthetic databases of species traits.
One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis; that is, analyzing relationships among species when estimating the effects of species traits within a community. 
Although there are statistical tools available and widely used for this type of problem, these tools are either inflexible, or computationally inefficient and unpractical, or both when fitting large volumes of data or complex models.
In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic relationships via mixed effect models, highlighting various complexities of phylogenetic relations through random effects.

In the last few decides, researcher have constantly develop new phylogenetic comparative methods and have shown importance of including phylogenetic relationships in these type of problems \cite{felsenstein1985phylogenies, li2017statistical}. 
The core of the existing techniques are based on different modelling assumptions of the evolutionary process using various parametrizations of phylogenetic correlations (PC) of taxa; that is matrices inferred from the topology of evolutionary trees by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods. 
For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods assuming Brownian motion (pure genetic draft) process. 
Despite the wide usage of these methods, there are a handful of recognizable limitations with these approaches and noticeably many published examples in the literature are constrained to small balanced data sets and simple models; that is, simply looking at phylogenetic relationship among species via traits. \citep{li2017canfun}.\ml{add more, should be easy. I don't really like this line, skipping for now}

@@

The typical phylogenetic regression is of the form:
\begin{align}
y & = XB + \epsilon \\
\epsilon & \sim \textrm{MVN}(0,\sigma^{2}C),
\label{eq:gls}
\end{align}
where $y$ is an $n \times 1$ vector of measurable response; $X$ is an $n \times m$ model matrix containing 1's in the first column and the $m$ predictor variables (phenotypic traits or environmental variables); $B$ is an $m$ vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).

% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
%\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
An alternative modelling approach is to use linear mixed effects modelling framework \citep{lynch1991methods}.
The typical linear mixed model has the form:
\begin{align}
Y & \sim F(\mu) \\
g(\mu) & = XB + Zb + \epsilon_{mm} \\
\epsilon_{mm} & \sim \textrm{MVN}(0,\sigma^2)
\end{align}
where $Z$ is an $n \times m$ model matrix for the $m$ -- dimensional vector-valued $m$ predictor variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$, $Z$ to an $n \times 1$ matrix such that:

\begin{align}
Z & \sim \textrm{MVN}(0,\sigma^{2}C).
\end{align}

@@

An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
In particular, suppose that the evolutionary process is a Brownian motion, which means the evolution of a continuous trait is a random walk and daughter species after speciation are all independent.  
In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
Thus, the evolutionary history for each species can be modelled with a sequence of independent errors with species--branch matrix $S$, rather than having to impose a correlation $C$. 
This is mathematically equvilent under BM assumption where we can reconstruct C by matrix multiplying $S$ by itself.

@@

\begin{equation}
Z^{C}_{i} = (S^{T}_{i}J^{T}_{i} \ast X^{T}_{i})^{T},
\end{equation}

@@

We generated test data based on the simple mixed model formulation \ml{ref equation (3-5)} with a single response variable $Y$ and a single predictor continuous variable $X$ for $n$ (20, 100, and 500) species, and $m$ (1 and 20) sites. 
To account for multiple levels of phylogenetic relations, we assume phylogenetic relations in the intercept $b_0$, and slope $b_1$.
We further assume that random slope and intercept are correlated. 
For models with multiple sites, we also assume for species and site interactions random effects, an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.

@@

Our algorithmic approach is general and could be implemented in a wide range of computational platformsthat support independent latent variables, we implemented our approach using the lme4 R package \citep{bates2015fitting}.
We have included three additional R packages that can fit phylogenetic compariative models, and they are nlme \citep{pinheiro2014r}, phylolm \citep{ho2014phylolm}, and pez \citep{pearse2015pez}.
Phylogenetic generalized least squares (PGLS) in nlme, is one of the most widely used techniques in phylogenetic compariative analysis; it

@@

We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical pcm to simulated data. 
First, this new approach is magnitudes faster compared to existing phylogenetic mixed models without losing robustness/accuracy and can easily combined with any statistically mixed model framework. 
Last and most importantly, it is more flexible in fitting large phylogenies, unbalanced datasets, and complex random effects such as slope correlations.

>>>>

Ever-increasing capabilities of data collection techniques (i.e. genetic sequencing, geographic information systems monitoring, and etc) and sampling efforts from field biologist are making large volumes of ecological/biological data available over an ever-wider range of species and also higher resolution (i.e. measurable phenotypic traits to genetic sequence data). 
As a result, researchers are interested in exploring more complicated systems and fitting more models to data or synthetic databases of species traits.
One particular area of interest in ecological/evolutionary statistics is phylogenetic comparative analysis; that is, analyzing relationships (signals) among species when estimating the effects of species traits within a community. 
Although there are statistical tools available and widely used for this type of problem, these tools are either inflexible, or computationally inefficient and unpractical, or both when fitting large volumes of data or complex models.
In this paper, we propose an alternative method that can flexibly and computational efficiently model phylogenetic signal via mixed effect models, highlighting various complexities of phylogenetic signals through random effects.

In the last few decides, researcher have constantly develop new phylogenetic comparative methods and have shown importance of including phylogenetic signal in these type of problems \cite{felsenstein1985phylogenies, li2017statistical}. 
The core of the existing techniques are based on different modelling assumptions of the evolutionary process using various parametrizations of phylogenetic correlations (PC) of taxa; that is matrices inferred from the topology of evolutionary trees by capturing the evolutionary changes that occurred on all the branches, which can be incorporated in statistical methods. 
For example, phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}, phylogenetic generalized least squares (PGLS) \citep{grafen1989phylogenetic}, phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} are methods assuming Brownian motion (pure genetic draft) process. 
Despite the wide usage of these methods, there are a handful of recognizable limitations with these approaches and noticeably many published examples in the literature are constrained to small balanced data sets and simple models; that is, simply looking at phylogenetic signal among species via traits. \citep{li2017canfun, freckleton2002phylogenetic}.

@@

The typical phylogenetic regression is of the form:
\begin{align}
y & = XB + \epsilon \label{eq:gls1} \\ 
\epsilon & \sim \textrm{MVN}(0,\sigma^{2}C), \label{eq:gls2}
\end{align}
where $y$ is an $n \times 1$ vector of measurable response; $X$ is an $n \times m$ model matrix containing 1's in the first column and the $m$ predictor variables (phenotypic traits or environmental variables); $B$ is an $m$ vector of coefficients; $\epsilon$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma^{2}_{\epsilon}C$ in which $C$ is an $n \times n$ matrix of the phylogenetic correlation (PC).

% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
%\ml{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
An alternative modelling approach is to use linear mixed effects modelling framework \citep{lynch1991methods}.
The typical linear mixed model has the form:
\begin{align}
Y & \sim F(\mu) \label{eq:glmm1} \\
g(\mu) & = XB + Zb + \epsilon_{mm} \label{eq:glmm2} \\
\epsilon_{mm} & \sim \textrm{MVN}(0,\sigma^2) \label{eq:glmm3}
\end{align}
where $Z$ is an $n \times m$ model matrix for the $m$ -- dimensional vector-valued $m$ predictor variables; $\epsilon_{mm}$ is the residual error and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\sigma_{2}I$.
Analogously, the phylogenetic regression given by (1) can be represented in the mixed model framework by constraining $\epsilon_{mm} = 0$, $Z$ to an $n \times 1$ matrix such that:

\begin{align}
Z & \sim \textrm{MVN}(0,\sigma^{2}C). \label{eq:glmmgls}
\end{align}

@@

An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
In particular, suppose that the evolutionary process is a Brownian motion, which means the evolution of a continuous trait is a random walk and daughter species after speciation are all independent.  
In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $S$, rather than having to impose a correlation $C$. 
This is mathematically equivalent under BM assumption where we can reconstruct C by matrix multiplying $S$ by itself.

@@

\begin{equation}
Z^{C}_{i} = (S^{T}_{i}J^{T}_{i} \ast X^{T}_{i})^{T}, \label{eq:ZC}
\end{equation}

@@

We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $Y$ and a single predictor continuous variable $X$ for $n$ (20, 100, and 500) species, and $m$ (1 and 20) sites/groups. 
To account for multiple levels of phylogenetic signals, we assume phylogenetic signals in the intercept $b_0$, and slope $b_1$.
We further assume that random slope and intercept are correlated. 
For models with multiple sites, we also assume for species and site interactions random effects, an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.

@@

Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables, we implemented our approach using the lme4 R package \citep{bates2015fitting}.
We have included three additional R packages that can fit phylogenetic comparative models under BM process assumption, and they are nlme \citep{pinheiro2014r}, phylolm \citep{ho2014phylolm}, and pez \citep{pearse2015pez}.
Phylogenetic generalized least squares (PGLS) in \textit{nlme}, is one of the most widely used techniques in phylogenetic comparative analysis; it allows for the analysis of relationships among species traits in its residual variance and can also handle multiple predictor variables.
Phylogenetic generalized linear model (PGLM) in \textit{phylolm}, is a slightly more flexible variation of PGLS that can allow for within species variation (sometimes called the tip variation). 
Neither PGLS nor PGLM can fit random slope or multiple sites. 
One of the relatively few packages that currently fit random slopes or multiple sites is \textit{pez}, which can handle adding additional uncorrelated random slopes with intercept and species in multiple sites/groups.

@@

We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical pcm to simulated data. 
First, this new approach is magnitudes faster compared to existing phylogenetic mixed models without losing robustness/accuracy and can easily combined with any statistically mixed model framework. 
Last and most importantly, it is more flexible in fitting large phylogenies, unbalanced data sets, and complex random effects such as slope correlations.
